---
title: Biased Estimators - Ridge and Lasso regressions
author: Michele Scandola
date: '2020-08-22'
slug: biased-estimators-ridge-and-lasso-regressions
categories:
- Statistics
tags:
- (Generalized) Linear Models
- R
subtitle: ''
lastmod: '2020-08-22T17:30:13+02:00'
authorLink: 'https://michelescandola.netlify.app/'
description: ''
hiddenFromHomePage: yes
hiddenFromSearch: no
featuredImage: ''
featuredImagePreview: ''
toc:
  enable: yes
math:
  enable: yes
lightgallery: no
license: ''
---

```{r}
library(knitr)
library(captioner)
fig_nums <- captioner()
cod_nums <- captioner( prefix = "Code" )
```


Some days ago I wrote some notes concerning
[unbiased estimators](/presentations/20200819linearregression/unbiased-linear-regression/index.html),
that are the estimators that we commonly use with linear models, and statistical
analyses such as ANOVA, linear regression, etc...

_Unbiased_ estimators try to estimate the best unknown but not casual,
$\beta$ coefficients of the _independent variables_ (i.v.)
to explain the observed _dependent variable_ $y$ (d.v.),
minimising the error between the variables that can be sampled from the 
linear model ($\hat{y}$) and the observed data ($y$).

In other terms, they try to reduce the error between $y$ and $\hat{y}$ with
different methodologies. We can also say that they look for the values
of the coefficients that better "explain the variance".

## Problems with the Unbiased Estimators

That's great! However, having models that predict so brilliantly the behaviour
of a specific sample of data coming from a statistical population, can lead
to the problem that the model is not actually so useful in predicting new data
and, therefore, it has not really captured the values of the coefficients
at a _population-level_. This is called _overfitting_.

> [_Overfitting_](https://en.wikipedia.org/wiki/Overfitting):
the production of an analysis that corresponds too
closely or exactly to a particular set of data, and may therefore fail to fit
additional data or predict future observations reliably

This may also be caused by the use of an excessive number of i.v. in the
analysis.

In order to prevent this problem, _biased_ estimators have been proposed.

> _Biased estimators_: algorithms to estimate the values of the coefficients
with penalisation coefficients, in order to have coefficients that do not
allow the perfect correspondence between $\hat{y}$ and $y$.

Concepts that are often used to understand better when a statistical model
is a good model are the _bias_ and the _variance_. Please, note that in this
context _bias_ has
no relation with cognitive bias, and _variance_ in this context is not the
dispersion statistics that we all know.

> The _bias_ is the error of the model in fitting the observed data $y$.
A large _bias_ means that the model _underfits_ the sample of data.

```{r}
x = 1:10
y = x^3/50 + rnorm(10, sd = 1)

plot(x,y)
abline(lm(y~x), col="red")
points(x , x^3/50, type="l", col="green")
legend("top", col = c("red" , "green"), lty = 1,
       legend = c("Underfit" , "Correct fit"))
```

`r fig_nums("underfit_plot",
             "Graphical representation of an hypothetical distribution fitted by two models, one with correct fit (y ~ x^3), and one that underfits the data (y ~ x)")`

> The _variance_ is the error of the model in predicting new data.
A large _variance_ means that the model _overfits_ the
original sample of data $y$.

```{r}
x = seq(from = 1, to = 5, by = 0.1)
y = x^3/100 + cos(x*10)^2

x1 = seq(from = 1, to = 5, by = 0.001)
y1 = x1^3/100 + cos(x1*10)^2

plot(x,y)
points(x1, y1, type="l",  col =" red")
points(x, x^3/100 + mean(cos(x*10)^2), type = "l", col = "green")
legend("topleft", col = c("red" , "green"), lty = 1,
       legend = c("Overfit" , "Correct fit"))


```

`r fig_nums("overfit_plot",
             "Graphical representation of an hypothetical distribution fitted by two models, one with correct fit (y ~ x^3), and one that overfits the data (y ~ x^3 + cos(x * 10)^2)")`
             
(to be honest, the overfitted curve is obtained by using exactly
y ~ x^3 + cos(x * 10)^2, but for simplicity let's pretend that the
cosine part is just random variation)

A good model should have low _bias_ and low _variance_, finding the trade-off between them.

```{r}
x = 1:100
y = x^5 / 1e10

plot( x = 1:100 , y , type = "l" , col = "red")
points( x = 100:1 , y = y , type = "l", col = "blue")
points( x = 1:100 , y = y + y[100:1] + 0.1 , type = "l" , col = "green")
legend("top", col = c("red" , "blue" , "green"), lty = 1, legend = c("Bias" , "Variance" , "Error trade-Off"))
```

`r fig_nums("bias_variace_tradeoff",
             "Graphical representation of the Bias and Variance errors of the model, and hypothetical trade-off curve.
             Please note that the Trade-off curve has been displaced for a better visualization")`



The most famous among the _biased estimators_ are the:

- Ridge regression (a.k.a. L2 regularization)
- Lasso regression (a.k.a. L1 regularization)

## Ridge regression

Ridge regression tries to find the best coefficients ($\beta$s) that describe
our observed sample of data ($y$), penalising the computation of the
coefficients with a coefficient $\lambda$.

Therefore, the idea is to find the $\beta$s of the model
that best fit the data, with an additional
penalysing term $\lambda$.

$\beta = \underset{\beta}{\operatorname{arg min}}~S(\beta, \lambda)$

where

$S(\beta, \lambda) = \sum_{i=1}^{n} \frac{1}{2} ( y_i - \sum_{j=1}^p x_{i,j}  \beta_{j})^2+\lambda \sum_{j=1}^p \beta_j^2$ 

$\lambda$ is the *complexity* parameter, and penalises the $\beta$s, limiting
*overfitting* and shrinking large $\beta$s.

In order to better understand how ridge regression works,
we can use an iterative, inefficient, and so simplified to be simply wrong,
function just to have an idea concerning how ridge regression works.

In this function we create some random data gaussianally distributed,
coming from a linear model with mean 1 ($\beta_0$) and 
regressor $\beta_1 = 2$.

The same data used for [unbiased estimators](/presentations/20200819linearregression/unbiased-linear-regression/index.html).

We set our $\lambda$ to 0.1.

```{r, fig.height=10, fig.width=10}
## creating the data
set.seed( 1 )
# independent variable
x <- 1:10
#   beta0 + i.v. * beta1 + epsilon
y <- 1 + x * 2 + rnorm( n = 10 )

find_ridge_least_squares <- function( y , x , beta0 , beta1 , lambda ) {
  
  # variable for the beta1 that best fits the data
  best_beta <- c(NULL, NULL)
  # variable for memorizing the squared errors
  best_err  <- Inf
  for(int in beta0){
    for( bet in beta1 ){
      yhat <- int + x * bet
      err  <- ( sum( ( y - yhat)^2 ) / 2 ) + ( lambda * ( int^2 + bet^2 ) )
      best_beta[1] <- ifelse( best_err > err,
                           int,
                           best_beta[1])
      best_beta[2] <- ifelse( best_err > err,
                           bet,
                           best_beta[2])
      best_err  <- ifelse( best_err > err,
                           err,
                           best_err)
    }
  }
  
  return( best_beta )
}

print(
  find_ridge_least_squares( y ,
                            x,
                            beta0 = seq(from = 0, to = 3, by = 0.01),
                            beta1 = seq(from = 1, to = 3, by = 0.01),
                            lambda = 0.01 )
)
```

`r cod_nums("ridge", "Exemplification code for a ridge regression")`

The estimations are not so far from the real values. However, let's
see how the real function works, in library `glmnet`:

```{r}
library("glmnet")

fit <- glmnet(y = y, x = cbind(1,x),
              lambda = 0.01, alpha = 0)

print( coef( fit ) )
```

`r cod_nums("glmnet_ridge", "Use of glmnet for ridge regression")`

The minimisation problem has a unique solution:

$\beta = (X^TX + \lambda I)^{-1}X^Ty$

```{r}
X = cbind( 1 , x )

solve( t(X) %*% X + 0.01 * diag(2) ) %*% t(X) %*% y
```

`r cod_nums("matrix_glmnet", "Matrix computation for Ridge regression")`

In this case we can see that the algorithms used in the `glmnet` function
and in the matrix computation are different.

`glmnet` is a complex function that allows to compute Ridge regression (alpha = 0),
Lasso regression (alpha = 1) and Elasticnet regression (alpha $\in$ [0, 1]),
and probably it is using
algorithms that are more advanced than the classical ones that I am using here.

## Lasso regression

The Least Absolute Shrinkage and Selection Operator (Lasso) regression
is very similar to the Ridge regression.

In fact, it uses the $\lambda$ coefficient of penalisation already seen
in the Ridge regression.

Therefore, the minimisation problem is always:

$\beta = \underset{\beta}{\operatorname{arg min}}~S(\beta, \lambda)$

but the $S$ function changes:

$S(\beta, \lambda) = \frac{1}{n}\sum_{i=1}^{n} ( y_i - \sum_{j=1}^p x_{i,j}  \beta_{j})^2+\frac{\lambda}{2} \sum_{j=1}^p |\beta_j|$ 

```{r}
find_lasso_least_squares <- function( y , x , beta0 , beta1 , lambda ) {
  
  # variable for the beta1 that best fits the data
  best_beta <- c(NULL, NULL)
  # variable for memorizing the squared errors
  best_err  <- Inf
  for(int in beta0){
    for( bet in beta1 ){
      yhat <- int + x * bet
      err  <- ( sum( ( y - yhat)^2 ) / ( length(yhat) ) ) +
        ( lambda / 2 * abs( int + bet ) )
      best_beta[1] <- ifelse( best_err > err,
                           int,
                           best_beta[1])
      best_beta[2] <- ifelse( best_err > err,
                           bet,
                           best_beta[2])
      best_err  <- ifelse( best_err > err,
                           err,
                           best_err)
    }
  }
  
  return( best_beta )
}

print(
  find_lasso_least_squares( y ,
                            x,
                            beta0 = seq(from = 0, to = 3, by = 0.01),
                            beta1 = seq(from = 1, to = 3, by = 0.01),
                            lambda = 0.01 )
)
```

`r cod_nums("Lasso", "Exemplification code for a Lasso regression")`

```{r}
fit <- glmnet(y = y, x = cbind(1,x),
              lambda = 0.01)

print( coef( fit ) )
```

`r cod_nums("glmnet_lasso", "Use of glmnet for Lasso regression")`

The minimisation problem can be divided in two steps:

1. finding the $\beta$s with OLS ($\beta = (X^TX)^{-1}X^Ty$)
2. applying the Lasso penalisation ($(1+N \frac{\lambda}{2})^{-1}$)

$\beta = (1+N \frac{\lambda}{2})^{-1}[(X^TX)^{-1}X^Ty]$

```{r}
X = cbind( 1 , x )

(1 + nrow(X) * 0.01 / 2)^(-1) *
solve( t(X) %*% X ) %*% t(X) %*% y
```

Also in this case the results are different from the `glmnet` function.
I suspect that in the `glmnet` function they are using more advanced versions
of the Lasso regression.
